# Reproducible Research: Peer Assessment 1

Please set the working directory to the repository root with `setwd()` before sourcing this file.

Setting knitr options.
``` {r knitr_options}
require(knitr)
opts_chunk$set(fig.path="figures/")
```

Load required libraries.
```{r load_libraries}
require(dplyr)
```

## Loading and preprocessing the data

The project makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day.
The data consists of two months of data from an anonymous individual collected
during the months of October and November, 2012 and include the number of steps
taken in 5 minute intervals each day.

The variables included in the dataset activity.zip are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total
of 17,568 observations in this dataset.

We load the dataset.
```{r load_data}
activity <- read.csv(unz("activity.zip", "activity.csv"))
```

We create a dataset without NAs.
```{r activity_no_NA}
activity_no_NA <- activity[complete.cases(activity),]
```

## What is mean total number of steps taken per day?

We create a dataset with the total number of steps taken per day.
```{r steps_per_day}
steps_per_day <- activity_no_NA %.% group_by(date) %.% summarise(total_steps = sum(steps))
```

We make a histogram of the total number of steps taken each day.
```{r hist_steps_per_day}
with(steps_per_day, hist(total_steps, main = "Total steps per day", xlab = ""))
```

We calculate and report the mean and median total number of steps taken per day.
```{r steps_mean_median}
with(steps_per_day, summary(total_steps))
```

## What is the average daily activity pattern?

We create a dataset with the average number of steps taken per 5-minute interval,
averaged across all days (y-axis)
```{r average_daily_activity}
average_daily_activity <- activity_no_NA %.% group_by(interval) %.% summarise(average_activity = mean(steps))
```

We make a time series plot of the 5-minute interval (x-axis) and the average
number of steps taken, averaged across all days (y-axis)
```{r plot_average_daily_activity}
with(average_daily_activity, plot(interval, average_activity, type = "l", main = "Average daily activity pattern", xlab = "Interval", ylab = "Number of steps"))
```

Which 5-minute interval, on average across all the days in the dataset, contains
the maximum number of steps?
```{r interval_with_max_activity}
interval_with_max_activity <- average_daily_activity %.% arrange(desc(average_activity)) %.% head(1)
interval_with_max_activity
```
The 5-minute interval, on average across all the days in the dataset, with the
maximum number of steps is `r interval_with_max_activity$interval`.

## Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA).
The presence of missing days may introduce bias into some calculations or summaries
of the data.

We calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)
```{r total_number_missing}
total_number_missing <- sum(!complete.cases(activity), na.rm=TRUE)
total_number_missing
```
The total number of missing values in the dataset is `r total_number_missing`.

We create a new dataset that is equal to the original dataset but with the missing data filled in
with average steps for that 5-minute interval.
```{r fill_missing_data}
activity_with_mean_steps_added <- left_join(activity, average_daily_activity)
activity_corrected <- activity_with_mean_steps_added[,c(1,2,3)]
for (i in 1:nrow(activity_corrected))
  if (is.na(activity_corrected[i,1])) activity_corrected[i,1] <- activity_with_mean_steps_added[i,4]
```

We create a dataset with the average number of steps taken per 5-minute interval,
averaged across all days (y-axis), using the original dataset with missing data filled.
```{r steps_per_day_corrected}
steps_per_day_corrected <- activity_corrected %.% group_by(date) %.% summarise(total_steps = sum(steps))
```

We make a histogram of the total number of steps taken each day, using the original
dataset with missing data filled.
```{r hist_steps_per_day_corrected}
with(steps_per_day_corrected, hist(total_steps, main = "Total steps per day", xlab = ""))
```

We calculate and report the mean and median total number of steps taken per day,
using the original dataset with missing data filled.
```{r steps_mean_median_corrected}
with(steps_per_day_corrected, summary(total_steps))
```

These values doesn't differ from the estimates from the first part of the project.
It seems that excluding missing values or filling them doesn't affect the estimates.

## Are there differences in activity patterns between weekdays and weekends?

We create a new factor variable in the dataset with two levels - "weekday" and "weekend"
indicating whether a given date is a weekday or weekend day.
```{r add_weekdays}
for (i in 1:nrow(activity_corrected))
  if (any(c('Mon','Tue','Wed','Thu','Fri') == weekdays(as.Date(activity_corrected[i,2]), abbreviate = TRUE))) {
    activity_corrected[i,4] <- "weekday"    
  } else {
    activity_corrected[i,4] <- "weekend"    
  }
activity_corrected[,4] <- as.factor(activity_corrected[,4])
names(activity_corrected)[4] <- "weekday"
```

We create a dataset with the average number of steps taken per weekday and 5-minute interval,
averaged across all days (y-axis). Then we create others two dataset with rows filtered
for weekdays and weekend days.
```{r average_daily_activity_weekdays}
average_daily_activity_corrected <- activity_corrected %.% group_by(weekday, interval) %.% summarise(average_activity = mean(steps))
average_daily_activity_weekday <- average_daily_activity_corrected[average_daily_activity_corrected$weekday == "weekday",]
average_daily_activity_weekend <- average_daily_activity_corrected[average_daily_activity_corrected$weekday == "weekend",]
```

We make a panel plot containing a time series plot of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r plot_average_daily_activity_weekdays, fig.height=10.5, fig.width=8}
par(mfrow=c(2,1),oma=c(5,5,5,0))
par(mar=c(2, 3, 3, 2))
with(average_daily_activity_weekend, plot(interval, average_activity, type = "l", main = "Weekend", xlab = "", ylab = ""))
par(mar=c(3, 3, 2, 2))
with(average_daily_activity_weekday, plot(interval, average_activity, type = "l", main = "Weekday", xlab = "", ylab = ""))
mtext("Average daily activity pattern", outer = TRUE, side = 3) 
mtext("Number of steps", outer = TRUE, side = 2)
mtext("Interval", outer = TRUE, side = 1) 
```
