PA2 Exploration
===============

## Preparation

Setting knitr options.
``` {r knitr_options}
require(knitr)
opts_chunk$set(fig.path="figures/ex-")
```

Load required libraries.
```{r load_libraries, warning=FALSE}
require(dplyr)
require(lubridate)
require(quantmod)
require(ggplot2)
require(xtable)
require(choroplethr)
```

## Load and munge storm data

```{r read_storm_data, cache=TRUE}
storm_data_raw <- read.csv(bzfile("StormData.csv.bz2"))
```

```{r munge_storm_data, cache=TRUE}
# Function to decode the EXP symbol
decode_exp <- function(exp_symbol) {
  # Normalize to millions of dollars
  if (toupper(exp_symbol) == "B") exp <-1000
  else if (toupper(exp_symbol) == "M") exp <-1
  else if (toupper(exp_symbol) == "K") exp <- 1/1000
  else if (toupper(exp_symbol) == "H") exp <- 1/10000
  # Don't know how to interpet other values
  else exp <-0
  return(exp)
}
decode_exp_v <- Vectorize(decode_exp)
# Conversion factor to calculate amounts in constant dollars
getSymbols("CPIAUCSL", src='FRED') # Get Consumer Price Index from Federal Reserve Economic Data
avg.cpi <- apply.yearly(CPIAUCSL, mean) # Calculate an annual average
cf <- 1 / avg.cpi * as.numeric(avg.cpi['2011']) # Calculate conversion factor using 2011 as the base year
# Munge data
storm_data <- storm_data_raw %.%
  mutate(property_damage_exp = decode_exp_v(PROPDMGEXP)
        ,crop_damage_exp = decode_exp_v(CROPDMGEXP)
        ,event_year = year(mdy_hms(BGN_DATE))
        ,factor_cd = as.numeric(cf[as.character(event_year)])
        ,property_damage = PROPDMG * property_damage_exp * factor_cd
        ,crop_damage = CROPDMG * crop_damage_exp * factor_cd) %.%
  select(event_year
        ,event_type = EVTYPE
        ,state = STATE
        ,fatalities = FATALITIES
        ,injuries = INJURIES
        ,property_damage
        ,crop_damage)
```

## Event Type Correction

The NWS documentation about the data reports only 48 event type. The numbers
of event types present in the files are in the 900th.
We could not correct all of them and assign one of the 48 correct ones.
So we analyzed the first 50 event type by count and apply corrections only to
them.

```{r event_type_correction, results='asis'}
storm_data_corrected <- storm_data
storm_data_corrected$event_type <- toupper(storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^(SMALL )?HAIL.*", "HAIL", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("TSTM|THUNDERSTORMS?", "THUNDERSTORM", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("STORMS?", "STORM", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("WINDS?|WINDS?/HAIL", "WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("RAINS?", "RAIN", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^TH?UN?DEE?RS?TO?RO?M ?WIND.*|^(SEVERE )?THUNDERSTORM$|^WIND STORM$|^(DRY )?MI[CR][CR]OBURST.*|^THUNDERSTORMW$", "THUNDERSTORM WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^COASTAL ?STORM$|^MARINE ACCIDENT$", "MARINE THUNDERSTORM WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^FLOODS?.*|^URBAN/SML STREAM FLD$|^(RIVER|TIDAL|MAJOR|URBAN|MINOR|ICE JAM|RIVER AND STREAM|URBAN/SMALL STREAM)? FLOOD(ING)?S?$|^HIGH WATER$|^URBAN AND SMALL STREAM FLOODIN$|^DROWNING$|^DAM BREAK$", "FLOOD", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^FLASH FLOOD.*|^RAPIDLY RISING WATER$", "FLASH FLOOD", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("WATERSPOUTS?", "WATERSPOUT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("WEATHER/MIX", "WEATHER", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("CURRENTS?", "CURRENT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^WINDCHILL$|^COLD.*|^LOW TEMPERATURE$|^UNSEASONABLY COLD$", "COLD/WIND CHILL", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^EXTREME WIND ?CHILL$|^(EXTENDED|EXTREME|RECORD)? COLDS?$", "EXTREME COLD/WIND CHILL", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^WILD/FOREST FIRE$|^(WILD|BRUSH|FOREST)? ?FIRES?$", "WILDFIRE", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^RAIN/SNOW$|^(BLOWING|HEAVY|EXCESSIVE|BLOWING|ICE AND|RECORD)? ?SNOWS?.*", "HEAVY SNOW", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^FOG$", "DENSE FOG", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^(GUSTY|NON-SEVERE|NON ?-?THUNDERSTORM)? ?WIND.*|^ICE/STRONG WIND$", "STRONG WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("SURGE$", "SURGE/TIDE", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("CLOUDS?", "CLOUD", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^FROST[/\\]FREEZE$|^FROST$|^(DAMAGING)? ?FREEZE$|^HYP[OE]R?THERMIA.*|^ICE$|^(ICY|ICE) ROADS$|^BLACK ICE$|^ICE ON ROAD$", "FROST/FREEZE", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^GLAZE.*|^FREEZING (RAIN|DRIZZLE|RAIN/SNOW|SPRAY$)$|^WINTRY MIX$|^MIXED PRECIP(ITATION)?$|^WINTER WEATHER MIX$|^LIGHT SNOW$|^FALLING SNOW/ICE$|^SLEET.*", "SLEET", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^HURRICANE.*", "HURRICANE/TYPHOON", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^HEAT WAVES?$|^UNSEASONABLY WARM$|^WARM WEATHER$", "HEAT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^(EXTREME|RECORD/EXCESSIVE|RECORD) HEAT$", "EXCESSIVE HEAT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^HEAVY SURF(/HIGH SURF)?.*$|^(ROUGH|HEAVY) SEAS?.*|^(ROUGH|ROGUE|HAZARDOUS) SURF.*|^HIGH WIND AND SEAS$|^HIGH SURF.*", "HIGH SURF", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^LAND(SLUMP|SLIDE)?S?$|^MUD ?SLIDES?$|^AVALANCH?E$", "AVALANCHE", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^UNSEASONABLY WARM AND DRY$|^DROUGHT.*|^HEAT WAVE DROUGHT$", "DROUGHT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^TORNADO.*", "TORNADO", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^TROPICAL STORM.*", "TROPICAL STORM", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^MARINE MISHAP$|^HIGH WIND/SEAS$", "MARINE HIGH WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^HIGH WIND.*", "HIGH WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^HIGH SEAS$", "MARINE STRONG WIND", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^RIP CURRENT.*", "RIP CURRENT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^WATERSPOUT.*", "WATERSPOUT", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^EXCESSIVE RAINFALL$|^RAIN.*|^TORRENTIAL RAINFALL$|^(HEAVY|HVY)? (RAIN|MIX|PRECIPITATION).*", "HEAVY RAIN", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^FOG.*", "FREEZING FOG", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^WINTER STORM.*", "WINTER STORM", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^THUNDERSNOW$|^ICE STORM.*", "ICE STORM", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("WAVES?|SWELLS?", "SURF", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^LIGHTNING.*", "LIGHTNING", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^WHIRLWIND$|^GUSTNADO$|^TORNDAO$", "TORNADO", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^COASTAL FLOOD.*", "COASTAL FLOOD", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^TYPHOON", "HURRICANE/TYPHOON", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^EROSION/CSTL FLOOD$|^COASTAL FLOOD/EROSION$|^COASTAL SURGE/TIDE$", "COASTAL FLOOD", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^ASTRONOMICAL HIGH TIDE$", "STORM SURGE/TIDE", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^(GROUND)? ?BLIZZARD.*$", "BLIZZARD", storm_data_corrected$event_type)
storm_data_corrected$event_type <- gsub("^DUST STORM.*$", "DUST STORM", storm_data_corrected$event_type)
```

## Exploration

#### Events Summary

```{r event_year_with_impact_plot, results='asis', fig.width=12}
event_year_with_impact <- storm_data_corrected %.%
  filter(fatalities != 0 || injuries != 0 || property_damage != 0 || crop_damage != 0) %.%
  group_by(event_year) %.%
  summarise(event_per_year = n()) %.%
  arrange(event_year)
qplot(event_year, event_per_year, data = event_year_with_impact, geom = "path")
```

The events in the database start in the year 1950 and end in November 2011.
In the earlier years of the database there are generally fewer events recorded,
most likely due to a lack of good records. More recent years should be considered
more complete.
We will limit our analysis to the last 10 years: 2001-2011.

```{r year_filter_2001_2011, cache=TRUE}
storm_data_2001_2011 <- storm_data_corrected[storm_data_corrected$event_year >= 2001,]
```

### Fatalities Summary

```{r fatalities_summary}
sum(storm_data_2001_2011$fatalities)
summary(storm_data_2001_2011$fatalities[storm_data_2001_2011$fatalities != 0])
boxplot(storm_data_2001_2011$fatalities[storm_data_2001_2011$fatalities != 0], log = "y")
```

```{r event_type_fatalities_summary, results='asis'}
event_type_fatalities_summary <- storm_data_2001_2011 %.%
  group_by(event_type) %.%
  summarise(fatalities_per_type = sum(fatalities), event_count = n()) %.%
  mutate(fatalities_per_event_count = fatalities_per_type / event_count) %.%
  arrange(desc(fatalities_per_type, event_count))
print(xtable(event_type_fatalities_summary[event_type_fatalities_summary$fatalities_per_type != 0,]), type="html")
```

```{r top10_event_type_fatalities_year, results='asis', fig.height=15, fig.width=10}
top10_event_type_fatalities <- event_type_fatalities_summary[1:10,]$event_type
top10_event_type_fatalities_year <- storm_data_2001_2011 %.%
  filter(event_type %in% top10_event_type_fatalities) %.%
  group_by(event_year, event_type) %.%
  summarise(fatalities_per_type_year = sum(fatalities)) %.%
  arrange(event_year, event_type)
qplot(event_year, fatalities_per_type_year, data = top10_event_type_fatalities_year, facets = event_type ~ ., geom = "path")
```

### Injuries Summary

```{r injuries_summary}
sum(storm_data_2001_2011$injuries)
summary(storm_data_2001_2011$injuries[storm_data_2001_2011$injuries != 0])
boxplot(storm_data_2001_2011$injuries[storm_data_2001_2011$injuries != 0], log = "y")
```

```{r event_type_injuries_summary, results='asis'}
event_type_injuries_summary <- storm_data_2001_2011 %.%
  group_by(event_type) %.%
  summarise(injuries_per_type = sum(injuries), event_count = n()) %.%
  mutate(injuries_per_event_count = injuries_per_type / event_count) %.%
  arrange(desc(injuries_per_type, event_count))
print(xtable(event_type_injuries_summary[event_type_injuries_summary$injuries_per_type != 0,]), type="html")
```

```{r top10_event_type_injuries_year, results='asis', fig.height=15, fig.width=10}
top10_event_type_injuries <- event_type_injuries_summary[1:10,]$event_type
top10_event_type_injuries_year <- storm_data_2001_2011 %.%
  filter(event_type %in% top10_event_type_injuries) %.%
  group_by(event_year, event_type) %.%
  summarise(injuries_per_type_year = sum(injuries)) %.%
  arrange(event_year, event_type)
qplot(event_year, injuries_per_type_year, data = top10_event_type_injuries_year, facets = event_type ~ ., geom = "path")
```

### Property Damage Summary

```{r property_damage_summary}
sum(storm_data_2001_2011$property_damage)
summary(storm_data_2001_2011$property_damage[storm_data_2001_2011$property_damage != 0])
boxplot(storm_data_2001_2011$property_damage[storm_data_2001_2011$property_damage != 0], log = "y")
```

```{r event_type_property_damage_summary, results='asis'}
event_type_property_damage_summary <- storm_data_2001_2011 %.%
  group_by(event_type) %.%
  summarise(property_damage_per_type = sum(property_damage), event_count = n()) %.%
  mutate(property_damage_per_event_count = property_damage_per_type / event_count) %.%
  arrange(desc(property_damage_per_type, event_count))
print(xtable(event_type_property_damage_summary[event_type_property_damage_summary$property_damage_per_type != 0,]), type="html")
```

```{r top10_event_type_property_damage_year, results='asis', fig.height=15, fig.width=10}
top10_event_type_property_damage <- event_type_property_damage_summary[1:10,]$event_type
top10_event_type_property_damage_year <- storm_data_2001_2011 %.%
  filter(event_type %in% top10_event_type_property_damage) %.%
  group_by(event_year, event_type) %.%
  summarise(property_damage_per_type_year = sum(property_damage)) %.%
  arrange(event_year, event_type)
qplot(event_year, property_damage_per_type_year, data = top10_event_type_property_damage_year, facets = event_type ~ ., geom = "path")
```

### Crop Damage Summary

```{r crop_damage_summary}
sum(storm_data_2001_2011$crop_damage)
summary(storm_data_2001_2011$crop_damage[storm_data_2001_2011$crop_damage != 0])
boxplot(storm_data_2001_2011$crop_damage[storm_data_2001_2011$crop_damage != 0], log = "y")
```

```{r event_type_crop_damage_summary, results='asis'}
event_type_crop_damage_summary <- storm_data_2001_2011 %.%
  group_by(event_type) %.%
  summarise(crop_damage_per_type = sum(crop_damage), event_count = n()) %.%
  mutate(crop_damage_per_event_count = crop_damage_per_type / event_count) %.%
  arrange(desc(crop_damage_per_type, event_count))
print(xtable(event_type_crop_damage_summary[event_type_crop_damage_summary$crop_damage_per_type != 0,]), type="html")
```

```{r top10_event_type_crop_damage_year, results='asis', fig.height=15, fig.width=10}
top10_event_type_crop_damage <- event_type_crop_damage_summary[1:10,]$event_type
top10_event_type_crop_damage_year <- storm_data_2001_2011 %.%
  filter(event_type %in% top10_event_type_crop_damage) %.%
  group_by(event_year, event_type) %.%
  summarise(crop_damage_per_type_year = sum(crop_damage)) %.%
  arrange(event_year, event_type)
qplot(event_year, crop_damage_per_type_year, data = top10_event_type_crop_damage_year, facets = event_type ~ ., geom = "path")
```

### Top 10 events intersection

```{r top10_event_intersection}
top10_event_fatalities <- intersect(top10_event_type_injuries,top10_event_type_fatalities)
top10_event_damage <- intersect(top10_event_type_crop_damage,top10_event_type_property_damage)
top10_event_fatalities
top10_event_damage
```

### State

#### Events summary per State

```{r state_event_summary, results='asis'}
state_event_summary <- storm_data_2001_2011 %.%
  group_by(state) %.%
  summarise(event_per_state = n()) %.%
  arrange(desc(event_per_state))
print(xtable(head(state_event_summary, n = 20)), type="html")
```

```{r state_event_choroplethr, results='asis', fig.width=12}
state_event_choroplethr <- state_event_summary %.%
  select(region = state, value = event_per_state)
choroplethr(state_event_choroplethr, "state", title="Events per state", num_buckets=4)
```

#### Fatalities summary per State

```{r state_fatalities_summary, results='asis'}
state_fatalities_summary <- storm_data_2001_2011 %.%
  group_by(state) %.%
  summarise(fatalities_per_state = sum(fatalities)) %.%
  arrange(desc(fatalities_per_state))
print(xtable(head(state_fatalities_summary, n = 20)), type="html")
```

```{r state_fatalities_choroplethr, results='asis', fig.width=12}
state_fatalities_choroplethr <- state_fatalities_summary %.%
  select(region = state, value = fatalities_per_state)
choroplethr(state_fatalities_choroplethr, "state", title="Fatalities per state", num_buckets=4)
```

#### Injuries summary per State

```{r state_injuries_summary, results='asis'}
state_injuries_summary <- storm_data_2001_2011 %.%
  group_by(state) %.%
  summarise(injuries_per_state = sum(injuries)) %.%
  arrange(desc(injuries_per_state))
print(xtable(head(state_injuries_summary, n = 20)), type="html")
```

```{r state_injuries_choroplethr, results='asis', fig.width=12}
state_injuries_choroplethr <- state_injuries_summary %.%
  select(region = state, value = injuries_per_state)
choroplethr(state_injuries_choroplethr, "state", title="Injuries per state", num_buckets=4)
```

#### Property damage summary per State

```{r state_pdamage_summary, results='asis'}
state_pdamage_summary <- storm_data_2001_2011 %.%
  group_by(state) %.%
  summarise(property_damage_per_state = sum(property_damage)) %.%
  arrange(desc(property_damage_per_state))
print(xtable(head(state_pdamage_summary, n = 20)), type="html")
```

```{r state_pdamage_choroplethr, results='asis', fig.width=12}
state_pdamage_choroplethr <- state_pdamage_summary %.%
  select(region = state, value = property_damage_per_state)
choroplethr(state_pdamage_choroplethr, "state", title="Property damage per state", num_buckets=4)
```

#### Crop damage summary per State

```{r state_cdamage_summary, results='asis'}
state_cdamage_summary <- storm_data_2001_2011 %.%
  group_by(state) %.%
  summarise(crop_damage_per_state = sum(crop_damage)) %.%
  arrange(desc(crop_damage_per_state))
print(xtable(head(state_cdamage_summary, n = 20)), type="html")
```

```{r state_cdamage_choroplethr, results='asis', fig.width=12}
state_cdamage_choroplethr <- state_cdamage_summary %.%
  select(region = state, value = crop_damage_per_state)
choroplethr(state_cdamage_choroplethr, "state", title="Crop damage per state", num_buckets=4)
```
